---
title: "Data tidying"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r warning=FALSE, message=FALSE}
library(dplyr)
library(tidyr)
library(readr)
```

# Data Sources

Original data are loaded from the KNB, from:

> Mike Byerly. Alaska commercial salmon catches by management region (1886- 1997). Gulf of Alaska Data Portal. df35b.304.2. https://knb.ecoinformatics.org/view/df35b.304.2

```{r}
catch_original <- read_csv("https://knb.ecoinformatics.org/knb/d1/mn/v2/object/df35b.302.1")
```


# Clean and Reshape data

## remove unnecessary columns

Drop All and Notes because we don't need them.

```{r}
catch_data <- catch_original %>% 
    select(Region, Year, Chinook, Coho, Pink, Chum, Sockeye)
head(catch_data)
```

## check column typing


```{r}
glimpse(catch_data)
```

- change chinook from character to numeric

```{r}
catch_clean <- catch_data %>% 
    mutate(Chinook = as.integer(Chinook))
```

- Figure out which rows have NAs
```{r}
i <- which(is.na(catch_clean$Chinook))
i
catch_original[i,]
```
- fix the row that has a character value and assign it the value `1`
```{r}
catch_clean <- catch_data %>% 
    mutate(Chinook = if_else(Chinook == "I", "1", Chinook)) %>% 
    mutate(Chinook = as.numeric(Chinook))
head(catch_clean)
```

## reshape data with Split-Apply-Combine

Switch from wide to long format...

```{r}
catch_long <- catch_clean %>% 
    pivot_longer(cols=c(-Region, -Year), names_to = "Species", values_to = "Catch") %>% 
    rename(catch_thousands = Catch) %>% 
    mutate(catch = catch_thousands*1000) %>% 
    select(-catch_thousands)

head(catch_long)
```

- And show how to pivot wider

```{r}
catch_wide <- catch_long %>% 
    pivot_wider(names_from = Region, values_from = catch)
head(catch_wide)
```

## Split-Apply-Combine

```{r}
mean_region <- catch_long %>% 
    filter(Species == "Sockeye" | Species == "Coho") %>% 
    group_by(Region, Species) %>% 
    summarize(catch_mean = mean(catch), n=n(), .groups = "drop")
head(mean_region)
```

# Join to Regions dataset

Load the regions table:

```{r}
region_defs <- read.csv("https://knb.ecoinformatics.org/knb/d1/mn/v2/object/df35b.303.1") %>% 
    select(code, mgmtArea) %>% 
    rename(Region = code)
head(region_defs)
```

```{r}
catch_joined <- left_join(catch_long, region_defs, by = "Region")
nrow(catch_joined) == nrow(catch_long)
head(catch_joined)
```

